Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2017-05-30 paper/submitted/round2_png/numerical_experiments_v4_cmame.tex
stash reviewer comments; start reviewer response doc and minor edits

The unknown parameters to be inferred correspond to the forcing field, so that in this case $f(q)=q$. The observations are sparse measurements of the state.
\Cref{fig:baseSetup} shows the locations of the observations and the region $\Omega_I$ over which the QoI is calculated. For the low-fidelity model, the inverse problem is linear (i.e., the observations depend linearly on the parameters). Since the inverse problem is ill-posed, we use Tikhonov regularization~\cite{EngHanNeu00}; the regularization term in the objective function \Cref{eq:invOpt_obj} is $R(q)=\frac{\beta}{2}\int_\Omega \|\nabla f(q)\|_2^2\:\textrm{d}A$, where $\beta=10^{-5}$ is the regularization coefficient.
%
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{baseSeries/setup_3_3.pdf}
\caption{Locations of the observations and the QoI region.}
\label{fig:baseSetup}
\end{figure}
%

@@

\Cref{tab:ref3D_diffmesh} shows the error at the end of each adaptive iteration. Each iteration of the adaptive algorithm uses the solution of the previous iteration as its initial guess. The number of degrees of freedom of each of the primary (and, if applicable, auxiliary) variables at each iteration is also given; the supplementary adjoint is solved on the high-fidelity mesh and thus each of its variables has the same number of degrees of freedom as each of the primary variables in the high-fidelity inverse problem. The high-fidelity inverse problem does not converge when the low-fidelity solution is used as an initial guess. Instead, we solve the inverse problem for an intermediate model
%
\begin{equation}
\nabla\cdot(n\vec{V}u - nD\nabla u) + k_ru^2 = f(q)
\end{equation}
%
and use the solution as an initial guess for the high-fidelity problem; equivalently, we use two steps of natural continuation on the reaction parameter: $k_r=0$, then $k_r=4.2\cdot10^{-4}$.
%
\begin{table}[htbp]
\label{tab:ref3D_diffmesh}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{DOFs} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative \\
& & & & (Estimated) & (Actual) & Error (Actual)  \\ \hline
LF   & 0    & 9537  & 168710 & -16463 & -85663 & -103    \\
MF01 & 5.2  & 13417 & 167366 & -7207  & -84319 & -102    \\
MF02 & 11.4 & 17895 & 89777  & -6208  & -6730  & -8.10   \\
MF03 & 16.3 & 21001 & 85880  & -2473  & -2833  & -3.41   \\
MF04 & 22.0 & 24528 & 83902  & -711   & -855   & -1.03   \\
MF05 & 27.7 & 27984 & 83119  & 32     & -72    & -0.087  \\
HF   & 100  & 70785 & 83047  & --     & --     & --    \\ \hline
\end{tabular}
\caption{Runtime and relative errors of adaptive algorithm iterations given relative error tolerance of $0.1\%$; relative errors are given with respect to the high-fidelity QoI estimate.}
\end{table}
%

@@

\Cref{tab:ref3D_newdata_QoI_diffmesh} shows the average QoI values, error estimates and solution times for the (non)linear solves. \Cref{tab:ref3D_newdata_times_diffmesh} shows the times needed to solve the inverse problems and to solve for the additional variables needed to obtain an error estimate. For six of the ten datasets, the high-fidelity inverse problem does not converge given the low-fidelity solution as an initial guess; these are solved using the $k_r=0$ solution as an initial guess so that true QoI errors can be calculated. The average high-fidelity inverse problem solution time shown in \Cref{tab:ref3D_newdata_times_diffmesh} includes only those cases (four of ten) that converged with the low-fidelity initial guess.
%
\begin{table}
\label{tab:ref3D_newdata_QoI_diffmesh}
\centering
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative  \\
& & & (Estimated) & (Actual) & Error (Actual) \\ \hline
LF   & 0    & 166774 & --    & -84326 & -102.3 \\
MF01 & 5.2  & 164597 & 4347  & -82149 & -99.65  \\
MF02 & 11.4 & 88867  & -5921 & -6418  & -7.79  \\
MF03 & 16.3 & 85237  & -2414 & -2789  & -3.38  \\
MF04 & 22.0 & 83411  & -724  & -963   & -1.17  \\
MF05 & 27.7 & 82664  & -18   & -216   & -0.26 \\
HF   & 100  & 82500  & --    & --     & --  \\ \hline
\end{tabular}
\caption{Average QoI values and errors from solving inverse problem with mixed- and high-fidelity models and noisy data; relative errors are with respect to true high-fidelity QoI.}
\end{table}

%
\begin{table}
\caption{Average times to solve inverse problem and obtain error estimate with mixed- and high-fidelity models and noisy datasets.}
\label{tab:ref3D_newdata_times_diffmesh}
\centering
\begin{tabular}{ccc|c|c|c}
\cline{4-5}
 & & & \multicolumn{2}{|c|}{Error Estimation} & \\
\cline{1-6}
\multicolumn{1}{|c|}{\multirow{3}{*}{Case}} & \multicolumn{1}{|c|}{\multirow{3}{*}{DOFs}} & Inverse & Auxiliary & Supplementary & \multicolumn{1}{|c|}{Total} \\
\multicolumn{1}{|c|}{} & \multicolumn{1}{|c|}{} & Problem & Variables & Adjoint & \multicolumn{1}{|c|}{Solution}\\
\multicolumn{1}{|c|}{} & \multicolumn{1}{|c|}{} & Time (s) &  Time (s) & Time (s) & \multicolumn{1}{|c|}{Time (s)}\\
\cline{1-6}
\multicolumn{1}{|c|}{LF}    & \multicolumn{1}{|c|}{9537}   & 16   & --  & -- & \multicolumn{1}{|c|}{--} \\ \hline
\multicolumn{1}{|c|}{MF01}  & \multicolumn{1}{|c|}{13147}  & 185  & 107 & 235 & \multicolumn{1}{|c|}{526} \\ \hline
\multicolumn{1}{|c|}{MF02}  & \multicolumn{1}{|c|}{17895}  & 328  & 169 & 206 & \multicolumn{1}{|c|}{703} \\ \hline
\multicolumn{1}{|c|}{MF03}  & \multicolumn{1}{|c|}{21001}  & 435  & 202 & 185 & \multicolumn{1}{|c|}{821} \\ \hline
\multicolumn{1}{|c|}{MF04}  & \multicolumn{1}{|c|}{24528}  & 406  & 201 & 188 & \multicolumn{1}{|c|}{795} \\ \hline
\multicolumn{1}{|c|}{MF05}  & \multicolumn{1}{|c|}{27984}  & 498  & 263 & 198 & \multicolumn{1}{|c|}{959} \\ \hline
\multicolumn{1}{|c|}{HF}    & \multicolumn{1}{|c|}{70785}  & 1185 & --  & --  & \multicolumn{1}{|c|}{1185} \\ \hline
\end{tabular}
\end{table}
%
We see that the mixed-fidelity models, when applied to noisy datasets different to those with which they were generated, continue to perform well in achieving a small error in the QoI while limiting the use of the high-fidelity model to less than a third of the domain. Given the same initial guess, the mixed-fidelity inverse problems takes less time on average than the high-fidelity inverse problem to solve, and they converge consistently.

>>>>

The unknown parameters to be inferred correspond to the forcing field, so that in this case $f(q)=q$. The observations are sparse measurements of the state.
\Cref{fig:baseSetup} shows the locations of the observations and the region $\Omega_I$ over which the QoI is calculated. For the low-fidelity model, the inverse problem is linear (i.e., the observations depend linearly on the parameters). Since the inverse problem is ill-posed, we use Tikhonov regularization~\cite{EngHanNeu00}. The regularization term in the objective function \Cref{eq:invOpt_obj} is $R(q)=\frac{\beta}{2}\int_\Omega \|\nabla f(q)\|_2^2\:\textrm{d}A$, where $\beta=10^{-5}$ is the regularization coefficient; in this work, $\beta$ is chosen arbitrarily, though methods exist for choosing optimal regularization coefficients~\cite{BelgeKilmerMiller02,Calvetal00}.
%
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{baseSeries/setup_3_3.pdf}
\caption{Locations of the observations and the QoI region.}
\label{fig:baseSetup}
\end{figure}
%

@@

\Cref{tab:ref3D_diffmesh} shows the error at the end of each adaptive iteration. Each iteration of the adaptive algorithm uses the solution of the previous iteration as its initial guess. The number of degrees of freedom of each of the primary (and, if applicable, auxiliary) variables at each iteration is also given; the supplementary adjoint is solved on the high-fidelity mesh and thus each of its variables has the same number of degrees of freedom as each of the primary variables in the high-fidelity inverse problem. The high-fidelity inverse problem does not converge when the low-fidelity solution is used as an initial guess. Instead, we solve the inverse problem for an intermediate model
%
\begin{equation}
\nabla\cdot(n\vec{V}u - nD\nabla u) + k_ru^2 = f(q)
\end{equation}
%
and use the solution as an initial guess for the high-fidelity problem; equivalently, we use two steps of natural continuation on the reaction parameter: $k_r=0$, then $k_r=4.2\cdot10^{-4}$.
%
\begin{table}[htbp]
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{DOFs} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative \\
& & & & (Estimated) & (Actual) & Error (Actual)  \\ \hline
LF   & 0    & 9537  & 168710 & -16463 & -85663 & -103    \\
MF01 & 5.2  & 13417 & 167366 & -7207  & -84319 & -102    \\
MF02 & 11.4 & 17895 & 89777  & -6208  & -6730  & -8.10   \\
MF03 & 16.3 & 21001 & 85880  & -2473  & -2833  & -3.41   \\
MF04 & 22.0 & 24528 & 83902  & -711   & -855   & -1.03   \\
MF05 & 27.7 & 27984 & 83119  & 32     & -72    & -0.087  \\
HF   & 100  & 70785 & 83047  & --     & --     & --    \\ \hline
\end{tabular}
\caption{Runtime and relative errors of adaptive algorithm iterations given relative error tolerance of $0.1\%$; relative errors are given with respect to the high-fidelity QoI estimate.}
\label{tab:ref3D_diffmesh}
\end{table}
%

@@

\Cref{tab:ref3D_newdata_QoI_diffmesh} shows the average QoI values, error estimates and solution times for the (non)linear solves. \Cref{tab:ref3D_newdata_times_diffmesh} shows the times needed to solve the inverse problems and to solve for the additional variables needed to obtain an error estimate. For six of the ten datasets, the high-fidelity inverse problem does not converge given the low-fidelity solution as an initial guess; these are solved using the $k_r=0$ solution as an initial guess so that true QoI errors can be calculated. The average high-fidelity inverse problem solution time shown in \Cref{tab:ref3D_newdata_times_diffmesh} includes only those cases (four of ten) that converged with the low-fidelity initial guess.
%
\begin{table}
\centering
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative  \\
& & & (Estimated) & (Actual) & Error (Actual) \\ \hline
LF   & 0    & 166774 & --    & -84326 & -102.3 \\
MF01 & 5.2  & 164597 & 4347  & -82149 & -99.65  \\
MF02 & 11.4 & 88867  & -5921 & -6418  & -7.79  \\
MF03 & 16.3 & 85237  & -2414 & -2789  & -3.38  \\
MF04 & 22.0 & 83411  & -724  & -963   & -1.17  \\
MF05 & 27.7 & 82664  & -18   & -216   & -0.26 \\
HF   & 100  & 82500  & --    & --     & --  \\ \hline
\end{tabular}
\caption{Average QoI values and errors from solving inverse problem with mixed- and high-fidelity models and noisy data; relative errors are with respect to true high-fidelity QoI.}
\label{tab:ref3D_newdata_QoI_diffmesh}
\end{table}

%
\begin{table}
\centering
\begin{tabular}{ccc|c|c|c}
\cline{4-5}
 & & & \multicolumn{2}{|c|}{Error Estimation} & \\
\cline{1-6}
\multicolumn{1}{|c|}{\multirow{3}{*}{Case}} & \multicolumn{1}{|c|}{\multirow{3}{*}{DOFs}} & Inverse & Auxiliary & Supplementary & \multicolumn{1}{|c|}{Total} \\
\multicolumn{1}{|c|}{} & \multicolumn{1}{|c|}{} & Problem & Variables & Adjoint & \multicolumn{1}{|c|}{Solution}\\
\multicolumn{1}{|c|}{} & \multicolumn{1}{|c|}{} & Time (s) &  Time (s) & Time (s) & \multicolumn{1}{|c|}{Time (s)}\\
\cline{1-6}
\multicolumn{1}{|c|}{LF}    & \multicolumn{1}{|c|}{9537}   & 16   & --  & -- & \multicolumn{1}{|c|}{--} \\ \hline
\multicolumn{1}{|c|}{MF01}  & \multicolumn{1}{|c|}{13147}  & 185  & 107 & 235 & \multicolumn{1}{|c|}{526} \\ \hline
\multicolumn{1}{|c|}{MF02}  & \multicolumn{1}{|c|}{17895}  & 328  & 169 & 206 & \multicolumn{1}{|c|}{703} \\ \hline
\multicolumn{1}{|c|}{MF03}  & \multicolumn{1}{|c|}{21001}  & 435  & 202 & 185 & \multicolumn{1}{|c|}{821} \\ \hline
\multicolumn{1}{|c|}{MF04}  & \multicolumn{1}{|c|}{24528}  & 406  & 201 & 188 & \multicolumn{1}{|c|}{795} \\ \hline
\multicolumn{1}{|c|}{MF05}  & \multicolumn{1}{|c|}{27984}  & 498  & 263 & 198 & \multicolumn{1}{|c|}{959} \\ \hline
\multicolumn{1}{|c|}{HF}    & \multicolumn{1}{|c|}{70785}  & 1185 & --  & --  & \multicolumn{1}{|c|}{1185} \\ \hline
\end{tabular}
\caption{Average times to solve inverse problem and obtain error estimate with mixed- and high-fidelity models and noisy datasets.}
\label{tab:ref3D_newdata_times_diffmesh}
\end{table}
%
We see that the mixed-fidelity models, when applied to noisy datasets different to those with which they were generated, continue to perform well in achieving a small error in the QoI while limiting the use of the high-fidelity model to less than a third of the domain. Given the same initial guess, the mixed-fidelity inverse problems takes less time on average than the high-fidelity inverse problem to solve, and they converge consistently.
